### 2020-06-23 manuscript/manuscript.rmd
revisions focused up until line 130, numbered figures

There has been considerable recent debate about the deepest split in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some previous analyses found support for the first split being between comb jellies and all other animals (the Ctenophora-sister hypothesis), while others found support the first split being between sponges and all other animals (Porifera-sister). These studies differed in a many respects, including the genes and species selected, molecular evolution models, and analysis software. Here we systematically explore this question under consistent conditions by synthesizing data and results from XX previous phylogenomic analyses and performing new standardized analyses. Our reanalyses are consistent with previous studies that recover Porifera-sister with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models. Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for phylogenetic methods development.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant animal relative. Since then many more studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Table 1). As the problem has become better characterized, it has become an interesting test-case for addressing difficult phylogenetic questions, which has attracted methodological interest well beyond investigators concerned with early animal evolution. These studies have taken a wide variety of approaches and results have varied across these different analyses. The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result. But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses. This variance is often confounded, for example when a particular model is only available in one software package.

To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized and systematic framework. Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Table 1). These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partition annotations.

![](figures/Figure_overview.png)
**Fig. XXFigure_overview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta (designated by the blue node). (D) A ctenophore. (E) A sponge.

@@

Models of molecular evolution describe how probable different evolutionary changes are, and can have a major impact on the topology of inferred phylogenies. Models have several components that each consider different aspects of the evolutionary process. The models used in the studies considered here differ according to three primary components: the exchangeability matrix $E$ , rates of evolution, and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the rate at which one amino acid changes to another (*e.g.* Poisson/F81, WAG or LG, and GTR). The rate of exchange can be further scaled, for example with a Gamma parameter that accommodates rate variation across sites. Heterogeneous equilibrium approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). Moreover, the Gamma-rate heterogeneity is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used. A detailed description of each component is provided in Supplementary Results.

High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1). Studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices. These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details). As a result, the genes selected for analysis varies widely between studies (Fig. XX). This reflects the different methods used to collect sequence data and different criteria for gene selection. There are several ways to summarize and assess this variation in gene sampling. We annotated each gene partition against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table 1). These annotations show XXX. We identified number of ribosomal RNA genes in each data matrix (Extended Data Fig. 2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig. 2D). We also identified BUSCO genes [@waterhouse2017busco] in each matrix (Extended Data Fig. 2A). The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are broadly shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco]. The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). We found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20% (Extended Data Fig. 2B). This is an indication that gene sampling criterial are not only based on breadth of sampling, but also on other criteria such as features of molecular evolution and sampling effects such as sifferences in expression level.

@@

**Fig. XXtaxon_rectangles.** An overview of previous studies. (A) A time-series plot to show the main conclusions of previous 13 phylogenomic studies that tested Ctenophora-sister (designated by the orange node) and Porifera-sister (designated by the green node). (B) Matrix taxa and gene sampling for each study. Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

Sampling of ingroup taxa (animals) has always been heavilly biased toward Bilateria, but sampling of non-bilaterian groups, including sponges and ctenophores, has improved over time as studies add additional taxa. As sequencing technologies have developed, there has also been some differences across studies in which members of particular groups are included as older shallower datasets are replaced with new deeper datasets from closely related species. Ingroup taxon sampling across clades varies widely between studies (figure_taxon_rectangles). The first studies that use new sequencing technologies tend to have a greater fraction of species belonging to Bilateria, since it includes many of the best-studied animal groups and all caononical model species where these technologies are first deployed. More than 50% taxa sampled were Bilateria in the earliest transcriptome (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and genome genome (Borrowiec2015 [@borowiec2015extracting]) studies considered here. Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig. XXXX).

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation neede]. To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].

@@

```
**Fig. XXfigure_support_analyses.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study. Each shape represents an phylogenetic analysis that has been analyzed by ML (designated by the circle) or BI inference (designated by the triangle). Analyses with a support of bootstrap values less than 90% or a posterior probability less than 0.95 are considered as unresolved. (A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (related to Supplementary Table XX). (B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (related to Supplementary Table XX).

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3A and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3A). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.

@@

Here we reanalyzed 17 matrices from 12 studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQ-TREE (Extended Data Table 1). We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQ-TREE strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig. 3B; Supplementary Table 3).

@@

ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")),nrow = 3,
          p3,p4, labels = c("C", "D"))
```
**Fig. XXsensitive.**  Sensitive analyses with representative data matrices analyzed by different number of substitutionary categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrcies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values are from the sensitive analyses are shown in Supplementary table XX.

@@

We retrieved matrices from each publication (Table 1), storing the raw data in this manuscript's version control repository. We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks. All changes made are tracked with git.

@@

#### Phylogenetic analyses in IQ-TREE
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g. site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing. For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

@@

**Extended Data Table 1.** TTable 1. Overview of data matrices used in this study.
```{r table_study_summary}
tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
	kable(tableXXdistribution)

@@

## Extended Figures
```{r figure_gene_composition,, fig.height = 6, fig.width = 10}

@@

**Extended Data Fig. XXfigure_gene_composition.** Annotation and representation of BUSCO and ribosomal genes in each data matrix. (A). The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions. (B). The percentage BUSCO annotations in each matrix. (C). The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions. (D). The percentage ribosomal genes in each matrix.

**Extended Data Fig. XXfigure_matrix_overlap.** Pairwise overlap between each of the primary matrices considered here. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled. The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.

@@

# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
#                    widths = unit(c(4, 4), "cm"),
#                    heights = unit(c(2, 5), c("in", "lines")))
#
# grid.newpage()
# grid.draw(fig5)
```
**Extended Data Fig. XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices). The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with a log-likelihood difference of two as a gene with strong (lnl > 2) or weak (lnl < 2) phylogenetic signal for each hypothesis (Smith et al. 2020).

@@

**Extended Data Fig. XXrandom_recoding.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

@@

**Extended Data Fig. XXrecoding_pp.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Fig XXRecoidng_summary. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.

>>>>

There has been considerable recent debate about the placement of the root in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenomic analyses find support for a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others found support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in a many respects, including the genes and species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 13 previous phylogenomic analyses and performing new standardized analyses. Our reanalyses are consistent with previous studies that recover Porifera-sister with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models. Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for phylogenetic methods development.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative. Since then many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1). These studies have taken a wide variety of approaches and results have varied across these different analyses. The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result. But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses. This mthodological variance is often confounded, for example when a particular model is only available in one software package.

To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Extended Data Table 1). These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partition annotations.

![](figures/Figure_overview.png)
**Fig. 1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta (designated by the blue node). (D) A ctenophore. (E) A sponge.

@@

The models used in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empically estimated rates (WAG or LG), or all independently estimated from the data (GTR). The analyses considered here have site homogenous exchangeability models, with the same matrix used for all sites in all genes. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homongenous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices. Model effect is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that acccomodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readilly assemble matrices with hundreds or thousands of protein-coding genes (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices. These take different approaches to sequence comparison, filtering, and orthology assessment. As a result, the genes selected for analysis varies widely between studies (Fig. 2, Extended Data Fig. 1). To explore this variation in gene sampling, we annotated each gene with against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table XX). These annotations show XXX. We identified the number of ribosomal protein genes in each data matrix (Extended Data Fig. 2A). We also identified BUSCO genes [@waterhouse2017busco] in each matrix (Extended Data Fig. 2B). The animal BUSCO dataset contains 978 genes that tend to occur as a single copy used to assess the completeness of genome assemblies [@waterhouse2017busco]. All matrices have than 50% of the BUSCO genes, and several less than 20%. The fact that so many BUSCO genes are not in these matrices reflects that the gene selection criteria include many additional factors, such as high expression (since so many phylogenetomic data are based on transcriptome data) and features of molecular evolution.

@@

**Fig. 2.** An overview of previous studies. (A) A time-series plot to show the main conclusions of previous 13 phylogenomic studies that tested Ctenophora-sister (designated by the orange node) and Porifera-sister (designated by the green node). (B) Matrix taxa and gene sampling for each study. Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

Ingroup taxon sampling across clades varies widely between studies (Fig. 2, Extended Data Fig. 1). Sampling of ingroup taxa (animals) was initially biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups. Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig. 1).

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation needes]. To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi).

@@

```
**Fig. 3.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study. Each shape represents an phylogenetic analysis that has been analyzed by ML (designated by the circle) or BI inference (designated by the triangle). Analyses with a support of bootstrap values less than 90% or a posterior probability less than 0.95 are considered as unresolved. (A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (related to Supplementary Table XX). (B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (related to Supplementary Table XX).

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3A and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Extended Data Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3A). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.

Inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].

@@

Here we reanalyzed 17 matrices from 13 studies that are used as main conclusions from their original paper (Extended Data Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQ-TREE (Extended Data Table 1). We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQ-TREE strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig. 3B; Supplementary Table 3).

@@

ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")),nrow = 3,
          p3,p4, labels = c("C", "D"))
```
**Fig. 4.**  Sensitive analyses with representative data matrices analyzed by different number of substitutionary categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrcies using the site-heterogeneous Poisson-CAT model with different categories” (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values are from the sensitive analyses are shown in Supplementary table XX.

@@

We retrieved matrices from each publication (Extended Data Table 1), storing the raw data in this manuscript's version control repository. We manually made some formatting changes to make the batch processing of the matrices work well, *e.g.* standardizing the format of Nexus CHARSET blocks. All changes made are tracked with git.

@@

#### Phylogenetic analyses in IQ-TREE
To investigate the phylogenetic hypotheses and distribution of phylogenetic signal in studies aiming to find the root position of animal phylogeny, we considered 16 data matrices from all phylogenomic studies that were constructed from EST, transcriptomic, or genomic data (Extended Data Table 1). Because different choices of substitution models could largely influence phylogenetic inference of the placement of the root position of animal phylogeny (e.g. site-heterogeneous vs. site-homogeneous models), we first investigated model-fit from each matrix using ModelFinder in IQ-TREE v1.6.7, including site-heterogenous C10 to C60 profile mixture models (C60 models) as variants of the CAT models in ML framework (C10-C60 model were included for model comparison via -madd option). We included models that are commonly used in previous analyses, including site-homogeneous poisson, WAG+G, LG+G, GTR+G models plus C10-C60 models in the model testing. For computational efficiency, the GTR+C60 models were not included in model testing since it requires to estimate over 10,000 parameters. For large matrices like those from Hejnol2009, Borrowiec2015, and Simion2017, model testing is also not computational feasible so only LG+C60 models were used since LG/WAG+C60 models were suggested as the best-fit model in all other matrices.

@@

**Extended Data Table 1.** Table 1. Overview of data matrices used in this study.
```{r table_study_summary}
tableXXdistribution = read_tsv("../data_processed/tables/study_summary.tsv")
	kable(tableXXdistribution)

@@

## Extended Figures

**Extended Data Fig. 1.** Pairwise overlap between each of the primary matrices considered here. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled. The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.

```{r figure_gene_composition,, fig.height = 6, fig.width = 10}

@@

**Extended Data Fig. 2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix. (A). The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions. (B). The percentage BUSCO annotations in each matrix. (C). The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions. (D). The percentage ribosomal genes in each matrix.

@@

# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
#                    widths = unit(c(4, 4), "cm"),
#                    heights = unit(c(2, 5), c("in", "lines")))
#
# grid.newpage()
# grid.draw(fig5)
```
**Extended Data Fig. 3** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanozoa and full matrices). The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green). By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with a log-likelihood difference of two as a gene with strong (lnl > 2) or weak (lnl < 2) phylogenetic signal for each hypothesis (Smith et al. 2020).

@@

**Extended Data Fig. 4.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

@@

**Extended Data Fig. 5.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Fig XXRecoidng_summary. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.
